*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-22 Code/Manuscript_O2.Rmd
working on introduction

*Prochlorococcus marinus*, a genus of Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999].  *P. marinus* growth is currently constrained between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

@@

Clades LLII and LLIII are grouped together as the second oldest phylogenetic lineages diversifying in the *Prochlorococcus* radiation, and share an affinity for low light; LLIII cultured strain SS120, was isolated from 135 m depth in the Sargasso Sea.

@@

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are  typically dominant picophytoplankton in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI, including cultured strain MED4,is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface [@partenskyProchlorococcusMarinePhotosynthetic1999].  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant group in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population [@partenskyProchlorococcusMarinePhotosynthetic1999],  and are most numerous around 50 m depth [@partenskyProchlorococcusMarinePhotosynthetic1999]. Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

*Prochlorococcus marinus* clades are found  in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes
[@westNichePartitioningProchlorococcusPopulationsStratified1999;
@zinserInfluenceLightTemperature2007;
@delmontLinkingPangenomesMetagenomes2018] 
[@moorePhotophysiologyMarineCyanobacterium1999]. 
@lavinNovelLineagesProchlorococcus2010 and
@mattisonPhotosyntheticStrategiesWild2020 show that LL ecotypes
can occupy regions in the OMZ at depths above 40 m, reaching ambient light levels above what LL clades were thought to tolerate.

@@

Our changing climate, characterized by warming ocean waters, is rapidly altering conditions for these highly specialized strains of marine picophytoplankton. Predictions indicate a net global increase of *P. marinus* cell abundances of 29%, along with  poleward latitudinal shifts of at least 10° in marine phytoplankton niches by the end of this century [@flombaumPresentFutureGlobal2013] in response to warming waters, with increases in *Prochlorococcus* of approximately 50% in the more poleward regions. As the solar incidence angles are lower and photoperiod seasonality is more pronounced at higher latitudes, this shift will carry these phytoplankters into new combinations of photoperiod, seasonal regimes and changing spectral regimes [@bartonAnthropogenicClimateChange2016]. A 10° poleward shift of *P. marinus*, coupled with a dramatic increase in cell abundance between 40°N and 50°N,  could have a large impact on ocean ecosystems and global biogeochemical cycles [@flombaumPresentFutureGlobal2013], and thus will undoubtedly impact the ecosystem balance off Canadian waters.

Global warming is also rapidly changing the chemistry of our oceans.  By the end of this century, surface ocean pH is projected to decline by 0.1 to 0.4  due to  projected increases in carbon dioxide concentrations [@garcia-sotoOverviewOceanClimate2021]. Moreover, substantial changes in the global water cycle, leading to extensive changes in worldwide precipitation patterns, are affecting ocean salinity levels on a global scale, and  ice melts due to rising temperatures are impacting salinity levels in the Arctic and Northwest Atlantic oceans [@leeRecognizingSalinityThreats2022]. Increasing sea temperatures are also causing a decrease in [O~2~] across our global oceans [@matearLongtermChangesDissolved2003], with significant declines in [O~2~] observed toward the poles in both hemispheres [@helmObservedDecreasesOxygen2011]. Warmer ocean waters increase stratification, and decrease oxygen solubility at the surface, which in turn decreases oxygen mixing downwards by ocean currents [@garcia-sotoOverviewOceanClimate2021]. Models predict that the OMZ in the Pacific and Indian Oceans are expanding [@garcia-sotoOverviewOceanClimate2021; @buseckeDivergingFatesPacific2022]. In light of these expanding OMZ, a recent study by @buseckeDivergingFatesPacific2022 highlights an intriguing finding; the cores of the OMZ, where the oxygen levels are lowest, is projected to contract.  Given the predicted decrease in overall oceanic [O~2~], projected expansion of OMZs, and possible decline in the OMZ core volumes, we set forth to examine the influence of [O~2~] on *P. marinus* growth, as this area of research has been relatively understudied.

>>>>

## *Prochlorococcus marinus* diversity
*Prochlorococcus marinus*, a genus of Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999].  *P. marinus* growth is currently constrained between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

@@

Clades LLII and LLIII are grouped together as the second oldest phylogenetic lineages diversifying in the *P. marinus* radiation, and share an affinity for low light; LLIII cultured strain SS120, was isolated from 135 m depth in the Sargasso Sea.

@@

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are typically dominant picophytoplankters in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI, including cultured strain MED4,is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface [@partenskyProchlorococcusMarinePhotosynthetic1999].  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant *P. marinus* clade in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population [@partenskyProchlorococcusMarinePhotosynthetic1999],  and are most numerous around 50 m depth [@partenskyProchlorococcusMarinePhotosynthetic1999]. Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

*P. marinus* clades are nonetheless found  in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes
[@westNichePartitioningProchlorococcusPopulationsStratified1999;
@zinserInfluenceLightTemperature2007;
@delmontLinkingPangenomesMetagenomes2018] 
[@moorePhotophysiologyMarineCyanobacterium1999]. 
@lavinNovelLineagesProchlorococcus2010, while LL ecotypes can occupy regions in OMZ at depths above 40 m, exploiting ambient light levels above what LL clades were thought to tolerate.

@@

Our changing climate, characterized by warming ocean waters, is rapidly altering conditions for these specialized clades of marine picophytoplankton. Predictions indicate a net global increase of *P. marinus* cell abundances of 29%, along with  poleward latitudinal shifts of at least 10° in marine phytoplankton niches by the end of this century [@flombaumPresentFutureGlobal2013] in response to warming waters, with increases in *P. marinus* of approximately 50% in the more poleward regions of their distributions. As solar incidence angles are lower, and photoperiod seasonality is more pronounced, at higher latitudes, polewards shifts will carry these phytoplankters into new combinations of photoperiod, seasonal regimes and changing spectral regimes [@bartonAnthropogenicClimateChange2016].

Global warming is also rapidly changing ocean chemistry.  By the end of this century, surface ocean pH is projected to decline by 0.1 to 0.4  due toprojected increases in carbon dioxide concentrations [@garcia-sotoOverviewOceanClimate2021]. Moreover, substantial changes in the global water cycle, leading to extensive changes in worldwide precipitation patterns, are affecting ocean salinity levels on a global scale, and ice melts due to rising temperatures are impacting salinity levels in the Arctic and Northwest Atlantic oceans [@leeRecognizingSalinityThreats2022]. Increasing sea temperatures are also causing a decrease in [O~2~] across our global oceans [@matearLongtermChangesDissolved2003], with significant declines in [O~2~] observed toward the poles in both hemispheres [@helmObservedDecreasesOxygen2011]. Warmer ocean waters increase stratification, and decrease oxygen solubility at the surface, which in turn decreases oxygen mixing downwards by ocean currents [@garcia-sotoOverviewOceanClimate2021]. Models predict that the OMZ in the Pacific and Indian Oceans are expanding [@garcia-sotoOverviewOceanClimate2021; @buseckeDivergingFatesPacific2022]. In light of these expanding OMZ, a recent study by @buseckeDivergingFatesPacific2022 highlights an intriguing finding; the cores of the OMZ, where the oxygen levels are lowest, is projected to contract.  Given the predicted decrease in overall oceanic [O~2~], projected expansion of OMZs, and possible decline in the OMZ core volumes, we set forth to examine the influence of [O~2~] on *P. marinus* growth, as this area of research has been relatively understudied.
